OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-05 paper/paper.Rmd
Compare OptiClust vs de novo vsearch
Start inserting numbers in results w/ R code

@@

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. 
There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
reference OTUs of an independent database, or 2) split the dataset into a
reference and query fraction, cluster the reference sequences _de novo_, then
fit the query sequences to the reference OTUs.
For each dataset repeated with 100 random seeds, we generated OTUs with OptiFit
using both strategies.
To compare to existing software, we also clustered OTUs _de novo_ using
OptiClust and VSEARCH, and with VSEARCH in reference-based mode against the
Greengenes database.
All clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the MCC as described previously [@westcott_opticlust_2017].
We calculated the fraction of query sequences that were fit to existing OTUs in
closed reference mode as an additional measure of quality for this mode.
_De novo_ clustering with OptiClust produced **X%** higher quality OTUs than
VSEARCH, but performed **Y%** slower than VSEARCH.

@@

@@

In open reference mode, fitting the datasets to reference OTUs with OptiFit
produced OTUs of similar quality as clustering the datasets _de novo_ with
OptiClust (**diff. in medians**).
This held true for all datasets and reference databases.
OptiFit produced higher quality OTUs than VSEARCH when open reference clustering
against the Greengenes database, with median MCC scores of **A** and **B%**
(respectively).
However, open reference VSEARCH and OptiClust respectively ran **X%** and **Y%**
faster than OptiFit in open reference mode.

>>>>

```{r deps, message = FALSE}
library(here)
library(mothuR)
library(tidyverse)
dat <- read_tsv(here('results', 'summarized.tsv'))
rel_diff <- function(final, init, percent = TRUE) {
  mult <- if (isTRUE(percent)) 100 else 1
  return((final - init) / init * mult)
}
```

@@

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. 
There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
reference OTUs of an independent database, or 2) split the dataset into a
reference and query fraction, cluster the reference sequences _de novo_, then
fit the query sequences to the reference OTUs.
For each dataset repeated with 100 random seeds, we generated OTUs with OptiFit
using both strategies.
To compare to existing software, we also clustered OTUs _de novo_ using
OptiClust and VSEARCH, and with VSEARCH in reference-based mode against the
Greengenes database.
All clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the MCC as described previously [@westcott_opticlust_2017].
We calculated the fraction of query sequences that were fit to existing OTUs in
closed reference mode as an additional measure of quality for this mode.

@@

```{r ref_db}
open_fit_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'mothur') %>% pull(mcc_median) %>% median()
opticlust_mcc <- dat %>% filter(method == 'de_novo', tool == 'mothur') %>% pull(mcc_median) %>% median()
open_fit_db_vs_clust <- rel_diff(opticlust_mcc, open_fit_mcc)

open_fit_gg_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'mothur', ref == 'gg') %>% pull(mcc_median) %>% median()
open_vsearch_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'vsearch', ref == 'gg') %>% pull(mcc_median) %>% median()
open_fit_db_vs_vsearch <- rel_diff(open_fit_gg_mcc, open_vsearch_mcc)
```

@@

In open reference mode, fitting the datasets to reference OTUs with OptiFit
produced OTUs of similar quality (`r open_fit_db_vs_clust`% difference in median
MCC) as clustering the datasets _de novo_ with OptiClust  across all datasets
and reference databases.
OptiFit produced higher quality OTUs than VSEARCH when open reference clustering
against the Greengenes database, with median MCC scores of `r open_fit_gg_mcc`
and `r open_vsearch_mcc` (respectively).
However, open reference VSEARCH and OptiClust respectively ran **X%** and **Y%**
faster than OptiFit in open reference mode.
_De novo_ clustering with OptiClust produced **X%** higher quality OTUs than
VSEARCH, but performed **Y%** slower than VSEARCH.
